---
title: "Field_Final"
author: "Jackson Sauls"
date: "2023-06-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up

First I will start by librarying my packages and importing my data. The data I will use for this 
project comes from Johns Hopkins University and describe trends in the Covid-19 pandemic over time.


```{r set up}
library(tidyverse)
library(lubridate)

link <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_US.csv","time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_US.csv","time_series_covid19_deaths_global.csv")

urls <- str_c(link,file_names)

global_cases <- read_csv(urls[2])
global_deaths <- read_csv(urls[4])
US_cases <- read_csv(urls[1])
US_deaths <-read_csv(urls[3])
```

## Tidy and Transform

Next I will tidy and transform the data to get it into a usable format.

```{r tidy}

global_cases <- global_cases %>%
  pivot_longer(cols = -c("Province/State","Country/Region",Lat,Long),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat,Long))

global_deaths <- global_deaths %>%
  pivot_longer(cols = -c("Province/State","Country/Region",Lat,Long),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat,Long))

global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = "Country/Region",
         Province_State = "Province/State") %>%
  mutate(date = mdy(date))
  
global <- global %>% filter(cases > 0)

US_cases <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat,Long_))

US_deaths <- US_deaths %>%
  pivot_longer(cols = -c(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat,Long_))

US <- US_cases %>%
  full_join(US_deaths)

global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat,Long_,Combined_Key,code3,iso2,iso3,Admin2))

global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(c(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key))
  
```

Next I will view a summary of my cleaned data to make sure it looks right.

```{r summary}
summary(global)

global

summary(US)
```

## Analysis

Now I will analyze and visualize my data.

```{r analysis}

# plot aggregate cases and deaths over time

ggplot(data = global, aes(x = date)) +
  geom_line(aes(y = cases), color = "steelblue") +
  geom_line(aes(y = deaths), color ="darkred")

# create counts of total cases and total deaths for the largest countries

counts <- global %>%
  filter(Population > 100000000) %>%
  group_by(Country_Region) %>%
  mutate(sum_deaths = sum(deaths), sum_cases = sum(cases))
  
# plot total cases for the largest countries
  
ggplot(data = counts, aes(x = Country_Region)) +
  geom_col(aes(y=sum_cases), color="steelblue")

# plot total deaths for the largest countries

ggplot(data = counts, aes(x = Country_Region)) +
  geom_col(aes(y=sum_deaths), color="darkred")

# calculate ratio of deaths to cases for each country

ratio <- counts %>%
  mutate(dc = sum_deaths / sum_cases)

# plot ratio of deaths to cases for each country

ggplot(data = ratio, aes(x = Country_Region)) +
  geom_col(aes(y=dc), color="darkgreen")

# create linear model of ratio of deaths to cases as a function of population

model <- lm(dc ~ Population, data = ratio)

summary(model)

```

## Conclusion

In conclusion, I analyzed the counts of cases and deaths for the top largest countries. I then calculated the ratio of deaths to cases and created a linear model that analyzed this ratio as a function of population. In the future, I would like to look further into this ratio to understand what factors to contribute to higher death ratio in some countries. My hypothesis would be that high poverty levels in some countries may be a factor. 


## Identification of Bias

One possible source of bias is that the data for this analysis is aggregated from many different sources. Each country has their own public health agency that gathers the data on Covid-19 for their country, which is then aggregated by Johns Hopkins. It is possible that some countries may try to deflate their number of cases or deaths to make themselves look better. 

A source of person bias is that I come from the U.S. and am not familiar with the culture of health
in other countries. The types of questions that I have about this data may be different than that of someone from another country. 


