---
title: "NYC Shooting Incident Data"
output:
  html_document: default
  pdf_document: default
  word_document: default
date: "2023-06-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Data

The data that I will use for this assignment comes from Data.gov. It is titled "NYPD Shooting Incident Data (Historic)" and contains records going back to 2006 on every shooting incident that occurred in NYC.

First I will load my libraries and import the data. 

```{r import}
library(tidyverse)
library(lubridate)

url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

nyData <- read_csv(url)
```

## Summary of Data

Next I will take a look at the first 10 rows and a summary of the data to get 
a better understanding of its structure. 

```{r summary}

nyData

summary(nyData)
```

## Tidy Data

Next I will tidy and transform my data to get it into a usable format. To do
this, I will set the OCCUR_DATE variable to the date datatype and rename the 
"Latitude", "Longitude", and "Lon_Lat" variables to match the all caps naming
convention of the other variables. I will also filter out the observations where
the latitude or longitude are set to "NA" because I plan to use these variables
later in my analysis.  I will then remove the variables that I will not need
for my analysis. Finally, I will create two new variables called "MONTH_YEAR" 
and "MONTH" that I will need for my analysis.

```{r tidy}
global <- nyData %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE)) %>%
  rename(LATITUDE = "Latitude", 
         LONGITUDE = "Longitude",
         LON_LAT = "Lon_Lat") %>%
  filter(is.na(LATITUDE) == FALSE,
          is.na(LONGITUDE) == FALSE) %>%
  select(-c(LOC_OF_OCCUR_DESC,PRECINCT,JURISDICTION_CODE,LOC_CLASSFCTN_DESC,
            LOCATION_DESC,STATISTICAL_MURDER_FLAG,X_COORD_CD,Y_COORD_CD,
            LON_LAT, INCIDENT_KEY, PERP_AGE_GROUP, PERP_SEX, PERP_RACE,
            VIC_AGE_GROUP, VIC_SEX, VIC_RACE))

global$MONTH_YEAR <- floor_date(global$OCCUR_DATE,"month")

global$MONTH <- substr(as.character(global$OCCUR_DATE),6,7)

summary(global)
```

## Analysis

### Location

First, I want to look at a count of the shootings by Boro to see if the shootings
are concentrated in one particular area.

```{r byboro}
BY_BORO <- global %>%
  group_by(BORO) %>%
  summarize(n = n())

ggplot(BY_BORO,aes(x =BORO,y=n)) +
  geom_col()
```



```{r}




```


### Time

I am also interested in looking at the total number of shootings over time to
see if there are any interesting trends. To make the graph easier to read, I 
will use the aggregated MONTH_YEAR variable that I created.

```{r overtime}

OVER_TIME <- global %>%
  group_by(MONTH_YEAR) %>%
  summarize(TOTAL = n())


ggplot(OVER_TIME,aes(x =MONTH_YEAR,TOTAL)) +
  geom_col()


```


From the chart above, I notice that the total number of shootings in a given month
tends to go up and down depending on the time of year. To look further into this,
I will look at the total number of shootings by month only.

```{r}
BY_MONTH <- global %>%
  group_by(MONTH) %>%
  summarize(TOTAL = n())


ggplot(BY_MONTH,aes(x =MONTH,y=TOTAL)) +
  geom_col()
```


Interestingly, it seems that shootings seem to increase around the summer months
and then decrease in the winter months, with slight increases around January and
February as well.


For my model, I will look at the median time of day of shootings as a function
of month. 


```{r}


mod <- global %>%
  group_by(MONTH) %>%
  summarize(MEDIAN_TIME = median(OCCUR_TIME))


ggplot(data = mod, aes(x = MONTH, y = MEDIAN_TIME)) +
         geom_col() +
         scale_y_time()
```


Based on the model, it looks like shootings tend to happen earlier in the day
during summer months and later in the day during winter months.


## Bias Identification

As a white man who is not from New York, it is possible that my background could
be a source of bias. I am familiar with stereotypes that crime tends to be higher
in high poverty neighborhoods, and much data shows that people of color are 
disproportionately targeted by police. 

One thing I found interesting about this data set is that the demographic information
was missing for a large number of perpetrators. It made me wonder how this data
is collected - does it come from police reports? I wonder if it is possible that 
this could introduce a source of bias as well.

It is important to be aware of bias when conducting data analysis, particularly
when it comes to sensitive issues such as gun violence. Gun violence is a topic that is often in the news, and unfortunately, many headline stories about gun violence 
and rising crime rates do not rely on the most accurate data. To mitigate my personal
bias, I tried to led the data lead my analysis rather than going into the process
with preconceived notions about what I might find.
